Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items

### 2025-06-13 analysis/primary_analyses.qmd
cite software versions
compute loading correlation for models with and without correlated errors

@@

```{r}
load_df <- 
  bind_rows(
    standardizedSolution(fit1A6P, type = "std.all") |> mutate(measure = "age", categories = "6"),
    standardizedSolution(fit1A2P, type = "std.all") |> mutate(measure = "age", categories = "2"),
    standardizedSolution(fit1G6P, type = "std.all") |> mutate(measure = "height", categories = "6"),
    standardizedSolution(fit1G2P, type = "std.all") |> mutate(measure = "height", categories = "2"),
    standardizedSolution(fit1Ge6P, type = "std.all") |> mutate(measure = "weight", categories = "6"),
    standardizedSolution(fit1Ge2P, type = "std.all") |> mutate(measure = "weight", categories = "2")) |>
  filter(lhs == "f1" & op == "=~") |> 
  mutate(item_name = rhs) |>
  # Create new columns: prefix (everything before underscore) and suffix (the numeric part)
  mutate(prefix = sub("_.*", "", item_name),
    suffix = as.numeric(sub(".*_", "", item_name))) |>
  group_by(prefix) |>
  mutate(item = match(suffix, sort(unique(suffix)))) |>
  mutate(item = replace_na(as.character(item), "Physical item")) |>
  ungroup() |>
  mutate(est_print = papaja::printnum(est.std, gt1 = FALSE, digits = 3)) |>
  select(item, measure, categories, est_print) |>
  pivot_wider(names_from = c(measure, categories), values_from = est_print)

@@

table2 <- rbind(
    load_df |> rename(Measure = item) |> mutate(group = "Items"),
    fit_df |> rename(Measure = statistic) |> mutate(group = "Fit")
  ) |> 
  group_by(group) |>
  flextable::as_grouped_data(groups = "group") |>
  flextable::as_flextable(hide_grouplabel = TRUE) |>
  flextable::add_header_row(values = c("Measure", "Age", "Height", "Weight"),
    colwidths = c(1, 2, 2, 2)) |>
  flextable::set_header_labels(
    values = list(
      Measure = "Categories",
      age_6 = "6", age_2 = "2", 
      height_6 = "6", height_2 = "2",
      weight_6 = "6", weight_2 = "2")) |>
  flextable::align(i = 1, align = "center", part = "header") |>
  flextable::border(i = 2, j = 1, border.top = officer::fp_border(width = 0), part = "header") |>
  flextable::border(i = 1, j = 1, border.bottom = officer::fp_border(width = 0), part = "header") |>
  flextable::theme_apa() |>
  ftExtra::colformat_md()
table2
```

@@

```{r}
results <- saveRDS(
  object = list(
    table1 = table1, table2 = table2, table3 = table3, table4 = table4
    ), 
  file = "../results/results.rds"
)
```

>>>>

```{r}
#| include: false
# record R packages
papaja::r_refs(file = "../r-references.bib")
```

@@

```{r}
load_df <- 
  bind_rows(
    standardizedSolution(fit1A6P, type = "std.all") |> mutate(measure = "age", categories = "6"),
    standardizedSolution(fit1A2P, type = "std.all") |> mutate(measure = "age", categories = "2"),
    standardizedSolution(fit1G6P, type = "std.all") |> mutate(measure = "height", categories = "6"),
    standardizedSolution(fit1G2P, type = "std.all") |> mutate(measure = "height", categories = "2"),
    standardizedSolution(fit1Ge6P, type = "std.all") |> mutate(measure = "weight", categories = "6"),
    standardizedSolution(fit1Ge2P, type = "std.all") |> mutate(measure = "weight", categories = "2")) |>
  filter(lhs == "f1" & op == "=~") |> 
  mutate(item_name = rhs)

load_df_proc <- load_df |>
  # Create new columns: prefix (everything before underscore) and suffix (the numeric part)
  mutate(prefix = sub("_.*", "", item_name),
    suffix = as.numeric(sub(".*_", "", item_name))) |>
  group_by(prefix) |>
  mutate(item = match(suffix, sort(unique(suffix)))) |>
  mutate(item = replace_na(as.character(item), "Physical item")) |>
  ungroup() |>
  mutate(est_print = papaja::printnum(est.std, gt1 = FALSE, digits = 3)) |>
  select(item, measure, categories, est_print) |>
  pivot_wider(names_from = c(measure, categories), values_from = est_print)

@@

table2 <- rbind(
    load_df_proc |> rename(Measure = item) |> mutate(group = "Items"),
    fit_df |> rename(Measure = statistic) |> mutate(group = "Fit")
  ) |> 
  group_by(group) |>
  flextable::as_grouped_data(groups = "group") |>
  flextable::as_flextable(hide_grouplabel = TRUE) |>
  flextable::add_header_row(values = c("Measure", "Age", "Height", "Weight"),
    colwidths = c(1, 2, 2, 2)) |>
  flextable::set_header_labels(
    values = list(
      Measure = "Categories",
      age_6 = "6", age_2 = "2", 
      height_6 = "6", height_2 = "2",
      weight_6 = "6", weight_2 = "2")) |>
  flextable::align(i = 1, align = "center", part = "header") |>
  flextable::border(i = 2, j = 1, border.top = officer::fp_border(width = 0), part = "header") |>
  flextable::border(i = 1, j = 1, border.bottom = officer::fp_border(width = 0), part = "header") |>
  flextable::theme_apa() |>
  ftExtra::colformat_md()
table2
```

@@

```{r}
results <- saveRDS(
  object = list(
    table1 = table1, table2 = table2, table3 = table3, table4 = table4,
    load_df_nomod = load_df
    ), 
  file = "../results/results.rds"
)
```

### 2025-06-13 analysis/secondary_analyses.qmd
cite software versions
compute loading correlation for models with and without correlated errors

```{r}
load_df <- 
  bind_rows(
    standardizedSolution(fit1A6, type = "std.all") |> mutate(measure = "age", categories = "6"),
    standardizedSolution(fit1A2, type = "std.all") |> mutate(measure = "age", categories = "2"),
    standardizedSolution(fit1G6, type = "std.all") |> mutate(measure = "height", categories = "6"),
    standardizedSolution(fit1G2, type = "std.all") |> mutate(measure = "height", categories = "2"),
    standardizedSolution(fit1Ge6, type = "std.all") |> mutate(measure = "weight", categories = "6"),
    standardizedSolution(fit1Ge2, type = "std.all") |> mutate(measure = "weight", categories = "2")) |>
  filter(lhs == "f1" & op == "=~") |> 
  mutate(item_name = rhs) |>
  # Create new columns: prefix (everything before underscore) and suffix (the numeric part)
  mutate(prefix = sub("_.*", "", item_name),
    suffix = as.numeric(sub(".*_", "", item_name))) |>
  group_by(prefix) |>
  mutate(item = match(suffix, sort(unique(suffix)))) |>
  mutate(item = replace_na(as.character(item), "Physical item")) |>
  ungroup() |>
  mutate(est_print = papaja::printnum(est.std, gt1 = FALSE, digits = 3)) |>
  select(item, measure, categories, est_print) |>
  pivot_wider(names_from = c(measure, categories), values_from = est_print)

@@

table5 <- rbind(
    load_df |> rename(Measure = item) |> mutate(group = "Items"),
    fit_df |> rename(Measure = statistic) |> mutate(group = "Fit")
  ) |> 
  group_by(group) |>
  flextable::as_grouped_data(groups = "group") |>
  flextable::as_flextable(hide_grouplabel = TRUE) |>
  flextable::add_header_row(values = c("Measure", "Age", "Height", "Weight"),
    colwidths = c(1, 2, 2, 2)) |>
  flextable::set_header_labels(
    values = list(
      Measure = "Categories",
      age_6 = "6", age_2 = "2", 
      height_6 = "6", height_2 = "2",
      weight_6 = "6", weight_2 = "2")) |>
  flextable::align(i = 1, align = "center", part = "header") |>
  flextable::border(i = 2, j = 1, border.top = officer::fp_border(width = 0), part = "header") |>
  flextable::border(i = 1, j = 1, border.bottom = officer::fp_border(width = 0), part = "header") |>
  flextable::theme_apa() |>
  ftExtra::colformat_md()
table5
```

@@

```{r}
results <- saveRDS(
  object = list(
    table5 = table5
    ), 
  file = "../results/results_secondary.rds"
)
```

>>>>

```{r}
load_df <- 
  bind_rows(
    standardizedSolution(fit1A6, type = "std.all") |> mutate(measure = "age", categories = "6"),
    standardizedSolution(fit1A2, type = "std.all") |> mutate(measure = "age", categories = "2"),
    standardizedSolution(fit1G6, type = "std.all") |> mutate(measure = "height", categories = "6"),
    standardizedSolution(fit1G2, type = "std.all") |> mutate(measure = "height", categories = "2"),
    standardizedSolution(fit1Ge6, type = "std.all") |> mutate(measure = "weight", categories = "6"),
    standardizedSolution(fit1Ge2, type = "std.all") |> mutate(measure = "weight", categories = "2")) |>
  filter(lhs == "f1" & op == "=~") |> 
  mutate(item_name = rhs)

load_df_proc <- load_df |>
  # Create new columns: prefix (everything before underscore) and suffix (the numeric part)
  mutate(prefix = sub("_.*", "", item_name),
    suffix = as.numeric(sub(".*_", "", item_name))) |>
  group_by(prefix) |>
  mutate(item = match(suffix, sort(unique(suffix)))) |>
  mutate(item = replace_na(as.character(item), "Physical item")) |>
  ungroup() |>
  mutate(est_print = papaja::printnum(est.std, gt1 = FALSE, digits = 3)) |>
  select(item, measure, categories, est_print) |>
  pivot_wider(names_from = c(measure, categories), values_from = est_print)

@@

table5 <- rbind(
    load_df_proc |> rename(Measure = item) |> mutate(group = "Items"),
    fit_df |> rename(Measure = statistic) |> mutate(group = "Fit")
  ) |> 
  group_by(group) |>
  flextable::as_grouped_data(groups = "group") |>
  flextable::as_flextable(hide_grouplabel = TRUE) |>
  flextable::add_header_row(values = c("Measure", "Age", "Height", "Weight"),
    colwidths = c(1, 2, 2, 2)) |>
  flextable::set_header_labels(
    values = list(
      Measure = "Categories",
      age_6 = "6", age_2 = "2", 
      height_6 = "6", height_2 = "2",
      weight_6 = "6", weight_2 = "2")) |>
  flextable::align(i = 1, align = "center", part = "header") |>
  flextable::border(i = 2, j = 1, border.top = officer::fp_border(width = 0), part = "header") |>
  flextable::border(i = 1, j = 1, border.bottom = officer::fp_border(width = 0), part = "header") |>
  flextable::theme_apa() |>
  ftExtra::colformat_md()
table5
```

@@

```{r}
results <- saveRDS(
  object = list(
    table5 = table5,
    load_df_mod = load_df
    ), 
  file = "../results/results_secondary.rds"
)
```

### 2025-06-13 manuscript.qmd
compute loading correlation for models with and without correlated errors

@@

Even after allowing for correlated errors as suggested by the modification indices, the loading patterns did not improve (see notes in @tbl-5). Loading patterns with and without modifications correlated above .98 across all questionnaires (body height, body weight, age) and response variants (two and six categories; *p* < .001). Therefore, no correlated errors were specified.

@@

Gold-standard items did not work as theoretically but empirically expected in this study. However, we still believe they are important. Only the gold standard item for body height – for six, not for two response categories – had its expected highest loading on the latent variable. @bortolottiRelevanceAdvantagesUsing2013, who previously applied the same procedure with two response categories to investigate a gold-standard item for body height, found a correlation of .86 with body height. This correlation pretty much matches our correlation with the latent variable. For our 2-response category scale, the correlation was slightly higher (.87), but there were also items which had descriptively equal or higher correlations (biserial correlations) with the sum score than the self-reported physical item (Pearson correlation). This can be due since we reported a latent correlation corrected for diattenuation, and our scale had 12 instead of 27 items. Nevertheless, our results depict a conceptual replication of @bortolottiRelevanceAdvantagesUsing2013 with a different questionnaire.

>>>>

```{r}
#| eval: false

# height
gold_loading_h6 <- results$load_df_nomod |>
  filter(categories == 6, measure == "height", rhs == "Koerpergroesse_realW") |>
  pull(est.std)
higher_load_h6 <- results$load_df_nomod |> 
  filter(categories == 6, str_detect(results$load_df_nomod$item_name, "KGr_")) |>
  summarize(higher_load = sum(est.std > gold_loading_h6))

gold_loading_h2 <- results$load_df_nomod |>
  filter(categories == 2, measure == "height", rhs == "Koerpergroesse_realW") |>
  pull(est.std)
higher_load_h2 <- results$load_df_nomod |> 
  filter(categories == 2, str_detect(results$load_df_nomod$item_name, "KGr_")) |>
  summarize(higher_load = sum(est.std > gold_loading_h2))

# weight
gold_loading_w6 <- results$load_df_nomod |>
  filter(categories == 6, measure == "weight", rhs == "Koerpergewicht_realW") |>
  pull(est.std)
higher_load_w6 <- results$load_df_nomod |> 
  filter(categories == 6, str_detect(results$load_df_nomod$item_name, "KGe_")) |>
  summarize(higher_load = sum(est.std > gold_loading_w6))

gold_loading_w2 <- results$load_df_nomod |>
  filter(categories == 2, measure == "weight", rhs == "Koerpergewicht_realW") |>
  pull(est.std)
higher_load_w2 <- results$load_df_nomod |> 
  filter(categories == 2, str_detect(results$load_df_nomod$item_name, "KGe_")) |>
  summarize(higher_load = sum(est.std > gold_loading_w2))

# age
gold_loading_a6 <- results$load_df_nomod |>
  filter(categories == 6, measure == "weight", rhs == "Alter_realW") |>
  pull(est.std)
higher_load_a6 <- results$load_df_nomod |> 
  filter(categories == 6, str_detect(results$load_df_nomod$item_name, "A_")) |>
  summarize(higher_load = sum(est.std > gold_loading_a6))

gold_loading_a2 <- results$load_df_nomod |>
  filter(categories == 2, measure == "age", rhs == "Alter_realW") |>
  pull(est.std)
higher_load_a2 <- results$load_df_nomod |> 
  filter(categories == 2, str_detect(results$load_df_nomod$item_name, "A_")) |>
  summarize(higher_load = sum(est.std > gold_loading_a2))
```

@@

```{r}
# correlation of loadings between cfa models with and without correlated errors
r_loadings <- cor.test(
  # use absolute values because the items were not recoded for cfa models
  abs(results$load_df_nomod$est.std), abs(results_secondary$load_df_mod$est.std), 
  method = "pearson")
r_loadings <- apa_print(r_loadings)$table
```

Even after allowing for correlated errors as suggested by the modification indices, the loading patterns did not improve (see notes in @tbl-5). Loading patterns with and without modifications correlated above `r r_loadings$estimate` across all questionnaires (body height, body weight, age) and response variants (two and six categories; *p* `r r_loadings$p.value`). Therefore, no correlated errors were specified.

@@

Gold-standard items did not work as theoretically but empirically expected in this study. However, we still believe they are important. Only the gold standard item for body height – for six, not for two response categories – had its expected highest loading on the latent variable. @bortolottiRelevanceAdvantagesUsing2013, who previously applied the same procedure with two response categories to investigate a gold-standard item for body height, found a correlation of .86 with body height. This correlation pretty much matches our correlation with the latent variable. For our 2-response category scale, the correlation was slightly higher (.87), but there were also items which had descriptively equal or higher correlations (biserial correlations) with the sum score than the self-reported physical item (Pearson correlation). This might be because we reported a latent correlation corrected for diattenuation, and our scale had 12 instead of 27 items. Nevertheless, our results depict a conceptual replication of @bortolottiRelevanceAdvantagesUsing2013 with a different questionnaire.
